How to Avoid OOTA Without Really Trying

### 2024-09-13 paper/ootavacuous.tex
ootavacuous: s/Single-Thread/Per-Thread/
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Apply second round September 13 Alan Stern feedback
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	We focus on shared-memory programs and C++ implementations based
	on traditional compilers and hardware, including CPUs and GPGPUs,
	thus enabling analysis of semantic dependencies and OOTA behaviors
	by means of existing hardware models.
	We show that these models' constraints prevent OOTA cycles from
	occurring in undefined-behavior-free C++ programs running on
	such implementations, provided the cycles involve only volatile
	atomics.
	We accommodate nonvolatile atomics by defining ``quasi-volatile''
	behavior that enables implementations doing per-thread analysis
	and optimization to not only avoid OOTA, but also to avoid 
	incorrectly evaluating arithmetic expressions involving atomic
	accesses.

@@

In broad terms, OOTA occurs when a group of threads load from each others'
stores and each thread's store depends on the value returned by that
thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air''
as shown in
Listing~\ref{lst:Simple OOTA}.
Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
values of 42, despite there being no instances of 42 in the default-zero
initial values or in the executable
code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
	Throughout this paper, variables whose names are ``\co{r}''
	followed by a number denote private per-process variables
	(``registers''),
	and other names denote shared variables.
	All shared-variable accesses are relaxed atomic.}
First, thread T0's line~1 does a load from \co{x} into \co{r1},
reading the value of T1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, T0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, T1's line~1 loads 42 from \co{y} to \co{r2}.
Finally, T1's line~2 stores 42 to \co{x}, justifying the value loaded
by T0's line~1.

@@

On the other hand, OOTA cycles need not involve nonsensical or
unmotivated values, as shown in
Listing~\ref{lst:OOTA Cycle}.
Line~3 of both processes stores the value 42 only when line~2 determines
that the value loaded was 42.
As with Listing~\ref{lst:Simple OOTA}, hardware ordering prevents the
OOTA result, as discussed further in
Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the Fundamental Property}.

@@

\subsection{For Compilers Using Single-Thread Analysis}
\label{sec:For Compilers Using Single-Thread Analysis}

>>>>

	We focus on shared-memory programs and C++ implementations based
	on traditional compilers and hardware, including CPUs and GPGPUs,
	thus enabling analysis of semantic dependencies and OOTA behaviors
	by means of existing hardware models.
	We show that these models' constraints prevent OOTA cycles from
	occurring in undefined-behavior-free C++ programs running on
	such implementations, provided the cycles involve only volatile
	atomics.
	We accommodate nonvolatile atomics by defining ``quasi-volatile''
	behavior that enables implementations doing per-thread analysis
	and optimization not only to avoid OOTA, but also to avoid 
	incorrectly evaluating arithmetic expressions involving atomic
	accesses.

@@

In broad terms, OOTA occurs when a group of threads load from each others'
stores and each thread's store depends on the value returned by that
thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air''
as shown in
Listing~\ref{lst:Simple OOTA}.
Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
values of 42, despite there being no instances of 42 in the default-zero
initial values or in the executable
code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
	Throughout this paper, variables whose names are ``\co{r}''
	followed by a number denote private per-thread variables
	(``registers''),
	and other names denote shared variables.
	All shared-variable accesses are relaxed atomic.}
First, thread T0's line~1 does a load from \co{x} into \co{r1},
reading the value of T1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, T0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, T1's line~1 loads 42 from \co{y} to \co{r2}.
Finally, T1's line~2 stores 42 to \co{x}, justifying the value loaded
by T0's line~1.

@@

On the other hand, OOTA cycles need not involve nonsensical or
unmotivated values, as shown in
Listing~\ref{lst:OOTA Cycle}.
Line~3 of both threads stores the value 42 only when line~2 determines
that the value loaded was 42.
As with Listing~\ref{lst:Simple OOTA}, hardware ordering prevents the
OOTA result, as discussed further in
Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the Fundamental Property}.

@@

\subsection{For Compilers Using Per-Thread Analysis}
\label{sec:For Compilers Using Per-Thread Analysis}
